---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!---
[![Travis-CI Build Status](https://travis-ci.org/.svg?branch=master)](https://travis-ci.org/)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/<USERNAME>/<REPO>?branch=master&svg=true)](https://ci.appveyor.com/project/<USERNAME>/<REPO>)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/polyggon)](https://cran.r-project.org/package=polyggon) 
[![Coverage Status](https://img.shields.io/codecov/c/github//master.svg)](https://codecov.io/github/?branch=master)
-->

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figure/README-"
)
library(tibble)
library(dplyr)
library(ggplot2)
```

## Polygons in R

```{r}
## taken from ?polypath
##"Nested rectangles, both clockwise")
library(tibble)
library(dplyr)
pts1 <- tibble(x = c(.1, .1, .9, .9, .2, .2, .8, .8),
              y = c(.1, .9, .9, .1, .2, .8, .8, .2), 
              group = c(1, 1, 1, 1, 2, 2, 2, 2), 
              id = 1)

# "Nested rectangles, outer clockwise, inner anti-clockwise")
pts2 <- tibble(x = c(.1, .1, .9, .9,  .2, .8, .8, .2),
               y = c(.1, .9, .9, .1,  .2, .2, .8, .8), 
              group = c(3, 3, 3, 3, 4, 4, 4, 4), 
              id = 2)
#               "Disjoint rectangles")
pts3 <- tibble(x = c(.1, .1, .4, .4, .6, .9, .9, .6),
               y = c(.1, .4, .4, .1,  .6, .6, .9, .9),
               group = c(5, 5, 5, 5, 6, 6, 6, 6), 
               id = 3)
#   "Overlapping rectangles, both clockwise")
pts4 <-   tibble(x = c(.1, .1, .6, .6, .4, .4, .9, .9),
               y = c(.1, .6, .6, .1, .4, .9, .9, .4),
               group = c(7, 7, 7, 7, 8, 8, 8, 8), 
               id = 4)
# "Overlapping rectangles, one clockwise, other anti-clockwise")
pts5 <- tibble(x = c(.1, .1, .6, .6, .4, .9, .9, .4),
               y = c(.1, .6, .6, .1, .4, .4, .9, .9),
               group = c(9, 9, 9, 9, 10, 10, 10, 10), 
               id = 5)
objects <- tibble(id = 1:5, 
                  description = c(
                    "Nested rectangles, both clockwise", 
                    "Nested rectangles, outer clockwise, inner anti-clockwise", 
                    "Disjoint rectangles", 
                    "Overlapping rectangles, both clockwise", 
                    "Overlapping rectangles, one clockwise, other anti-clockwise"
                  ))

          

pts <- bind_rows(pts1, pts2, pts3, pts4, pts5)
library(ggplot2)

## this doesn't look good
ggplot(inner_join(pts, objects)) + 
  aes(x = x, y = y, group = group, fill = factor(id)) + 
  geom_polygon() + facet_wrap(~ description, nrow = 5)
```

The problem is that `geom_polygon` uses `grid::polygonGrob` and this is not capable of drawing holes. 

It's as if we decided to use `polygon` in R, it's not going to work. Here I use transparency so we can see the overlapping polygons as slightly darker regions. 


```{r}
spreadpts <- pts %>% mutate(x = x + id, y = y + id)
split_insert_na <- function(x, f) {
  head(bind_rows(lapply(split(x, f), function(a) rbind(a, NA))), -1)
}

splitpts <- split(spreadpts, spreadpts$id)
op <- par(mfrow = c(5, 1), mar = rep(0.1, 4))
for (i in seq_along(splitpts)) {
  a <- splitpts[[i]]
  plot(a$x, a$y, axes = FALSE, xlab = "", ylab = "")
  polygon(split_insert_na(a, a$group), col = alpha("grey", 0.5))
}
par(op)
#lapply(split(spreadpts, spreadpts$id), function(a) polygon(split_insert_na(a, a$group), col = alpha("grey", 0.5)))
```

But, what if we use `polypath`?  Here I'm careful *not* to use transparency, as the behaviour is different on Windows for  `windows()` and `png()` - effectively the results is as if we used the `evenodd` rule no matter what `rule` is set to. 

```{r}
op <- par(mfrow = c(5, 1), mar = rep(0.1, 4))
for (i in seq_along(splitpts)) {
  a <- splitpts[[i]]
  plot(a$x, a$y, axes = FALSE, xlab = "", ylab = "")
  polypath(split_insert_na(a, a$group), col = "grey", rule = "winding")
  
}
par(op)
```

```{r}
op <- par(mfrow = c(5, 1), mar = rep(0.1, 4))
for (i in seq_along(splitpts)) {
  a <- splitpts[[i]]
  plot(a$x, a$y, axes = FALSE, xlab = "", ylab = "")
  polypath(split_insert_na(a, a$group), col = "grey", rule = "evenodd")
}
par(op)
```

## ggplot2?

There's no way to use `geom_polygon` to get these "polygons with hole" effects. We write a new `geom_holygon`, inspired by a post on the internet - but also include application across different `id` values as well as different group values. This is exactly analogous to the need to call `polypath` multiple times abvoe. 

```{r}
library(polyggon)
library(ggplot2)
library(grid)
descpoints <- inner_join(pts, objects) %>% mutate(border = 1)
ggplot(descpoints) + 
  aes(x = x, y = y, group = group, fill = factor(id), col = border) + 
  geom_holygon(rule = "winding") + facet_wrap(~ description, nrow = 5)



ggplot(descpoints) + 
  aes(x = x, y = y, group = group, fill = factor(id), col = border) + 
  geom_holygon(rule = "evenodd") + facet_wrap(~ description, nrow = 5)
```

But these are not very complicated!  

```{r}
library(rgdal)

iw <- readOGR(system.file("ext"))
